---
title: "Gov 50 Final Project"
author: "Kayra Yaman"
description: "Refuge vs Riches: The effect of GDP per capita on the number of refugees a country hosts per year"
output:
  distill::distill_article:
    self_contained: false
---
```{r Setup, echo= FALSE}
library(tidyverse)
library(infer)
library(modelsummary)
```


## Introduction
Does a country's GDP per capita influence the number of refugees/ internationally displaced people it hosts? In this study, I will determine whether refugees are more or less likely to be hosted by a wealthy, resource-rich country. This is a very timely question as refugee crises drag on across the world from the Syrian refugee crisis, that has left Turkey, a country with a struggling economy, hosting more than 3.6 million refugees, to the mass-migration border crisis affecting refugees and asylum seekers trying to reach the United States from South and Central America. I hypothesize that wealthier countries do not tend to host significantly more refugees than resource-poor countries. 


## Data
The data for this study is from two sources. Data on numbers of refugees, their origin countries, and their host countries (as stored in the "fled" tibble) is from the United Nations High Commissioner for Refugees' Refugee Data Finder (https://www.unhcr.org/refugee-statistics/download/?url=Y00Mtd). UNHCR collects this data primarily from governments around the world and from its own operations supporting refugees in those countries. Data on national, annual GDP per capita (as stored in the gdp tibble) is from the Organisation for Economic Co-operation and Development's Data Portal (https://data.oecd.org/gdp/gross-domestic-product-gdp.htm). The data is officially reported to the OECD from participating governments. The unit of analysis is a refugee, meaning that the outcome variable of interest is the the number of refugees arriving to a particular host country. It is important to note that from here on out the term "refugee" will be used to refer to all internationally displaced people, including official refugees accroding to the UNHCR, asylum-seekers, and other people in need of international protection. The explanatory variable of interest is a country's gross domestic product (or GDP) per capita accounting for purchasing power parity (PPP) rates and measured in US dollars. This variable was chosen to measure a country's wealth because it allows for a relatively accurate comparison of countries with different currencies and economies. Because this study uses data on the number of refugees who are hosted by more economically developed countries (treated) vs those hosted by less economically developed countries (control) collected from one point in time (a give year) to draw a comparison, it is a cross-sectional study. However, because the data is collected annually, it would be a repeated cross-sectional study, or one that collects data over time as well.


```{r Loading Data, echo= FALSE}
fled <- read_csv("data/population2.csv")

gdp <- read_csv("data/gdppercapita1.csv")
```


```{r Tidying UNHCR Data, echo= FALSE}

## I renamed the variables so that they would be easier to code with and then selected the renamed variables
fled_simple <- fled |>
  mutate(coo = `Country of origin`,
         coo_iso = `Country of origin (ISO)`,
         coa = `Country of asylum`,
         coa_iso = `Country of asylum (ISO)`,
         refugee = `Refugees under UNHCR's mandate`,
         asylumseek = `Asylum-seekers`,
         other = `Other people in need of international protection`,
         other = replace_na(other, 0),
         total = refugee + asylumseek + other) |>
  select(coo, coo_iso, coa, coa_iso, refugee, asylumseek, other, total, Year)

## I ungrouped the data by the country of origin in preparation to join it with the gdp tibble (to create a unique key)
merged <- fled_simple |>
  select(-c(coo, coo_iso)) |>
  group_by(coa_iso, Year) |>
  summarize(refugee_t = sum(refugee, na.rm = T),
            asylum_t = sum(asylumseek, na.rm = T),
            other_t = sum(other, na.rm = T),
            total_t = sum(total, na.rm = T))

```


```{r Tidying OECD Data, echo= FALSE}

## I filtered the data to only include entries measured by GDP per capita in USD, rather than total GDP in millions of USD, then selected relevant variables
gdp_simple <- gdp |>
  filter(MEASURE == "USD_CAP") |>
  rename(Year = TIME, coa_iso = LOCATION) |>
  select(Year, coa_iso, Value)


```


```{r Joining Tibbles, echo= FALSE}
## I identified unique keys for each tibble
gdp_check <- gdp_simple |>
  count(coa_iso, Year) |>
  filter(n > 1)

fled_check <- merged |>
  count(coa_iso, Year) |>
  filter(n > 1)

## Then I merged the tibbles using those keys
merged <- merged |>
  inner_join(gdp_simple, by = c("coa_iso", "Year"))

```


```{r Dependent Variable Plot, echo= FALSE}

dense_plot <- merged |>
  ggplot(aes(x = total_t)) +
  geom_density(color = "dodgerblue", size = 0.5) +
  labs(title = "Conitnuous Density Plot of Refugees Hosted Across the World",
       x = "Number of Refugees",
       y = "Density (country, year)")
dense_plot

```
The continuous density plot above visualizes the total number of refugees hosted per country per year. The plot is extremely right skewed with the point of highest density being near zero on the x-axis. This shows that the vast majority of countries in any year host very few (if not exactly zero) refugees.


## Results
```{r Independent and Dependent Variable Plot, echo= FALSE}
scatter_plot <- merged |>
  ggplot(aes(x = Value, y = refugee_t)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "dodgerblue") +
  labs(title = "The Effect of GDP on the Number of Refugees Hosted",
       subtitle = "by Country and by Year",
       x = "GDP per capita (in USD)",
       y = "Total Number of Refugees (by country by year")
scatter_plot

```
The scatter plot visualizes the relationship between a country's GDP per capita and the number of refugees it hosts per year. It shows a large cluster of countries per year with a GDP per capita of under US$50,000 and hosting 1 million refugees. The bivariate linear regression plotted over it has a slightly positive slope, suggesting that as a country's GDP per capita increases the number of refugees it hosts per year also increases slightly. However, in modelling this relationship, this bivariate regression does not control for year or country. 


```{r Linear Regression, echo= FALSE}
# g.lm <- glm(refugee_t ~ Value + Year + coa_iso, data = merged, family = poisson())
# summary(g.lm)

total.lm <- lm(total_t ~ Value + Year + coa_iso, data = merged)
c_total <- c("(Intercept)" = "Constant",
        "Value" = "GDP per capita (USD)",
        "Year" = "Year")
modelsummary(total.lm, coef_map = c_total, gof_map = c("nobs", "r.squared", "adj.r.squared"), stars = T) 
```
This multi-variable linear regression explains the relationship between a country's GDP and the number of refugees it hosts, controlling for year, and each country (though they are not shown in the above summary). This model has an R2 value of approximately 0.370 and an Adjusted R2 value of approximately 0.351 which indicates that the linear regression is not a particularly strong fit to the data. However, the regression coefficient of GDP shows that for an increase in $1 in GDP per capita, a given country would theoretically experience a decrease of approximately -2.652 in the total number of refugees it hosts, while controlling for country and year. This decrease is also statistically significant at the p < 0.01 level. Overall, this linear regression suggests that as a country's GDP per capita increases, the number of refugees it hosts decreases. 


## Conclusion
To conclude, while there does appear to be a statistically significant negative correlation between a country's GDP per capita and the number of refugees it hosts per year (when controlling for country and year), I cannot say that the relationship is definitively causal as there are many potential confounding factors that could also influence the number of refugees a country hosts, in addition to its GDP per capita. For example, geographic proximity to a conflict could be a confounding factor as many countries with higher GDPs per capita are not in close proximity to conflicts that might cause mass refugee migration and therefore might not receive as many refugees or asylum requests as the countries that are closer and that also have lower GDPs per capita. In addition, as mentioned above, the multi-variable linear regression constructed had fairly low R2 and Adjusted R2 values, indicating that the linear model was not a great fit for the data. This is one potential limitation of this study as a different type of regression might have resulted in a better model fit. Other potential limitations include the fact that the data on GDP and refugees hosted was collected from different sources, meaning that in joining the two data sets a number of rows had to be dropped (I used an inner join so rows were dropped from both tibbles). 

However, the statistically significant correlation, which shows that richer countries tend to host fewer refugees than poorer countries rings true to recent history. One important example is the famous EU-Turkey deal from 2016 (sometimes labeled as Europe's year of shame) when the European Union agreed to pay Turkey 6 billion euros to control its borders and prevent Syrian refugees from reaching the EU to apply for asylum and seek aid. This deal came about after Europe saw record numbers of refugees arriving to its shores the previous years, but that number was dwarfed by the 3.6 million Syrian refugees that Turkey (a country with a significantly smaller GDP per capita) currently hosts.


